Approximation as Differentiation

### 2025-07-10 main/language.tex
Improvements, not sure H(M) technically makes sense though.
Pass over section 5.
[ tex ] write up the semantics and some examples
[ tex ] fix bib

The syntax is defined in \figref{syntax}. Types includes base types $\rho$ drawn from $\PrimTy$, along with
standard type formers for sums, products, functions and lists, plus a type constructor $\tyLift$ to allow
approximation points to be added explicitly, as discussed in \secref{models-of-total-gps}. Terms include
variables, the usual introduction and elimination forms, a monadic return and bind for lifted terms, and
primitive operations $\phi$.

The language is intentionally minimal: it excludes general recursion, and general inductive or coinductive
types, which we will consider in future work (\secref{conclusion}). Typing judgments for terms are standard
and shown in \figref{typing}, with the usual rules for products, sums, functions, lists and lifting.

@@

A model of $\Sigma$ can be given in any category $\cat{C}$ with finite products, and assigns to each base type
$\rho \in \PrimTy$ an object $\sem{\rho}_{\PrimOp}$ in $\cat{C}$, and to each primitive operation $\phi \in
\PrimOp^\rho_{\rho_1,\ldots,\rho_n}$, a morphism $\sem{\phi}_{\Op}: \sem{\rho_1}_{\PrimTy} \times \ldots
\times \sem{\rho_n}_{\PrimTy} \to \sem{\rho}_{\PrimTy}$. A model of $\tyList$ can also be given in such a
category, and consists of an endofunctor $\List: \cat{C} \to \cat{C}$, plus for any object $X$, morphisms
$\nil: 1 \to \List(X)$ and $\cons: X \times \List(X) \to \List(X)$, and for any objects $Y, Z$ and morphisms
$f_\nil: Z \to Y$ and $f_\cons: Z \times X \times Y \to Y$, a morphism $\fold(f_\nil,f_\cons): Z \times
\List(X) \to Y$, with $\nil$, $\cons$ and $\fold$ satisfying the usual laws for lists.

The semantics for the full language is then given in terms of an arbitrary bicartesian closed category $\Sem$
providing models of $\Sigma$ and $\tyList$, as well as a strong monad $(\Lift, \eta, \mu)$ with strength
$\mathsf{st}$ for interpreting $\tyLift$. \figref{semantics:types} gives the interpretations of types $\tau$
and contexts $\Gamma$ as objects; \figref{semantics:terms} gives the interpretation of terms $\Gamma \vdash t:
\tau$ as morphisms $\sem{\Gamma} \to \sem{\tau}$.

\subsubsection{Interpretation for higher-order \GPS}

Given a model $M$ of $\Sigma$ in $\Fam(\LatGal)$, the category $\Fam(\MeetSLat \times \JoinSLat^\op)$ can
implement $\Sem$. It is bicartesian closed by \corref{mslat-jslat-bcc}, and can model $\tyList$ using infinite
coproducts (\secref{models-of-total-gps:coproducts-and-products}). The embedding $\HoEmbed: \Fam(\LatGal) \to
\Fam(\MeetSLat \times \JoinSLat^\op)$ from \propref{ho-embedding} preserves finite products, providing a model
$H(M)$ of $\Sigma$. Finally the lifting monad in $\Fam(\LatGal)$ (\secref{first-order:lifting}) is also
preserved by $H$. The correctness of this interpretation is considered in the next section.

>>>>

The syntax is defined in \figref{syntax}. Types includes base types $\rho$ drawn from $\PrimTy$, along with
standard type formers for sums, products, functions and lists% , plus a type constructor $\tyLift$ to allow
% approximation points to be added explicitly, as discussed in \secref{models-of-total-gps}
. Terms include
variables, the usual introduction and elimination forms% , a monadic return and bind for lifted terms
, and
primitive operations $\phi$.

The language is intentionally minimal: it excludes general recursion, and general inductive or coinductive
types, which we will consider in future work (\secref{conclusion}). Typing judgments for terms are standard
and shown in \figref{typing}, with the usual rules for products, sums, functions, and lists.

@@

An interpretation of a signature $\Sigma = (\PrimTy, \PrimOp)$ can be given in any category $\cat{C}$ with
finite products, and assigns to each base type $\rho \in \PrimTy$ an object $\sem{\rho}_{\PrimTy}$ in
$\cat{C}$, and to each primitive operation $\phi \in \PrimOp^\rho_{\rho_1,\ldots,\rho_n}$, a morphism
$\sem{\phi}_{\Op}: \sem{\rho_1}_{\PrimTy} \times \ldots \times \sem{\rho_n}_{\PrimTy} \to
\sem{\rho}_{\PrimTy}$.

Assuming that $\cat{C}$ is bicartesian closed and has a list object (\propref{lists}), then we can extend an
interpretation of a signature $\Sigma$ to an interpretation of the whole language over
$\Sigma$. \figref{semantics:types} and \figref{semantics:contexts} define the interpretation of types and
contexts as objects of $\cat{C}$ respectively. Terms are interpreted as morphisms between the interpretations
of the context and type, as defined in \figref{semantics:terms}. We have used the notations $\pi_i$ for
projections, $\prodM{f}{g}$ for pairing, $\coprodM{f}{g}$ for (parameterised) copairing, $!_x$ for morphisms
to the terminal object, and $\lambda$ and $\eval$ for currying and evaluation for exponentials.

For the first-order definability result in \secref{definability}, we will need another interpretation
$\sem{-}_{\mathit{fo}}$ of the first-order types (those constructed from primitive types, sums, unit and
products) in any bicartesian category. Such interpretations are preserved by finite coproduct and coproduct
preserving functors:

\begin{lemma}\label{lem:first-order-agreement-types}
  If $\cat{C}$ and $\cat{D}$ are bicartesian and bicartesian closed categories with interpretations of the
  signature $\Sigma$, $F : \cat{C} \to \cat{D}$ is a bicartesian functor, and
  $F(\sem{\rho}_{\PrimTy}) \cong \sem{\rho}_{\PrimTy}$ for all $\rho$, then for all first-order types $\tau$,
  $F(\sem{\tau}_{\mathit{fo}}) \cong \sem{\tau}$, and similar for contexts only containing first-order types.
\end{lemma}

\subsubsection{Interpretation for Higher-Order \GPS}

Given the above, we can now interpret the language in any of the bicartesian closed categories with list
objects we constructed in \secref{models-of-total-gps}. Specifically, we assume that we have an interpretation
of our chosen signature in $\Fam(\LatGal)$. Signatures are first-order, so it does not matter that
$\Fam(\LatGal)$ is not closed. Any such interpretation can be transported to
$\Fam(\MeetSLat \times \JoinSLat^\op)$ along the functor $H$ from \propref{ho-embedding} because it preserves
finite products. We can then interpret the whole language in $\Fam(\MeetSLat \times \JoinSLat^\op)$.

Interpreting a whole program $\Gamma \vdash t : \tau$ yield morphisms in
$\Fam(\MeetSLat \times \JoinSLat^\op)$ which, as in \secref{semantic-gps}, are triples
$(f, \partial f_f, \partial f_r)$ of the underlying function and the forward and backward approximation
maps. However, unlike in $\Fam(\LatGal)$ are not guaranteed that the forward and backward maps even operate on
the same lattices, let alone form a Galois connection. \lemref{first-order-agreement-types} guarantees that
the lattices agree, but the fact that the pair form a Galois connection is less trivial. We will prove this
property in \secref{definability}.

% Given a model of $\Sigma$ in $\Fam(\LatGal)$, the category $\Fam(\MeetSLat \times \JoinSLat^\op)$ can
% implement $\Sem$. It is bicartesian closed by \corref{mslat-jslat-bcc}, and can model $\tyList$ using infinite
% coproducts (\secref{models-of-total-gps:coproducts-and-products}). The embedding $\HoEmbed: \Fam(\LatGal) \to
% \Fam(\MeetSLat \times \JoinSLat^\op)$ from \propref{ho-embedding} preserves finite products, and so can
% transport each $\sem{\rho}_{\PrimTy}$ and $\sem{\phi}_{\Op}$ into the higher-order setting to provide a model
% of $\Sigma$. Finally the lifting monad in $\Fam(\LatGal)$ (\secref{first-order:lifting}) is also preserved by
% $H$. \roly{don't think we actually show this}

% The correctness of this interpretation is considered in the next section.

\subsection{Examples}

Let the signature
$\Sigma_{\mathsf{num}} = (\{\mathsf{num}\}, \{\mathrm{zero} : \tyUnit \to \mathsf{num}, \mathrm{add} :
\mathsf{num} \times \mathsf{num} \to \mathsf{num}\})$. This signature suffices to write the simple query
function in \exref{introduction-example}, where we interpret the $\mathrm{Label}$ type as the sum
$\tyUnit \tySum \tyUnit$ and the labels $\mathsf{a}$ and $\mathsf{b}$ as $\tmInl{\tmUnit}$ and
$\tmInr{\tmUnit}$. We consider several interpretations of $\Sigma_{\mathsf{num}}$ in $\Fam(\LatGal)$ and their
behaviour on the query from \exref{introduction-example}.
\begin{enumerate}
\item If we take $\sem{\mathsf{num}}_{\PrimTy} = \Disc(\mathbb{R})$, with $\sem{\mathrm{zero}}_{\PrimOp}$ and
  $\sem{\mathrm{add}}_{\PrimOp}$ the embeddings of the usual zero and addition functions via $\Disc$, then the
  resulting interpretation contains no approximation information. The approximation lattices for all types are
  always the trivial one-point lattice, and the forward and reverse approximation maps are necessarily the
  identity.
\item To get interesting approximation information, we take
  $\sem{\mathsf{num}}_{\PrimTy} = \Lift(\Disc(\mathbb{R}))$, using the lifting monad from
  \secref{models-of-total-gps:lifting-monad}, with $\sem{\mathrm{zero}}_{\PrimOp}$ and $\sem{\mathrm{add}}_{\PrimOp}$ defined from
  the unlifted interpretations above and the monad structure. This interpretation recovers the behaviour given
  in \exref{introduction-example}: running the reverse approximation map at the input $\mathsf{a}$ with the
  top approximation maps indeed reveals that only the numbers in the rows tagged with that label in the input
  are used, and likewise for $\mathsf{b}$.
\item More quantitative approximation information can be obtained by using the interval approximation
  interpretation from \exref{intervals-and-maxima-elements} and \secref{interval-approx}. We let
  $\sem{\mathsf{num}}_{\PrimTy} = \mathbb{R}_{\mathit{intv}}$ and interpret addition using the morphism given
  in \secref{interval-approx}.

  Recall that $\mathrm{query}\,\mathsf{a}\,\mathit{db} = 1$, so in the reverse direction we can pick an interval containing $1$ to discover the largest (i.e. least in the order) intervals that will give rise to this output as \emph{independent} changes to the input. For example, if we pick $[-0.9,1]$ as the interval, then:
  \begin{displaymath}
    (\partial\mathrm{query}\,\mathsf{a}\,\mathsf{db})_r([0.9,1.1]) = [(\cdot,[-0.1,0.1]),(\cdot,\bot),(\cdot,[0.9,1.1])]
  \end{displaymath}
  Indicating that, to achieve an output within $[0.9,1.1]$ for this query, either the first $\mathsf{a}$ row
  could be in $[-0.1,0.1]$ or the second one could be in $[0.9,1.1]$.
\end{enumerate}

\subsection{Systematic Insertion of Approximation via Moggi's CBN translation}
\label{sec:cbn-translation}

We can now carry out the systematic insertion of approximation points that we foreshadowed in
\secref{models-of-total-gps:lifting-monad}, using \citet[\S 3.1]{notions-of-computation}'s monadic CBN translation. We use the $T$
monad from \secref{tagging-monad} because it can be defined in terms of the language constructed we already
have. This requires that we have a signature $\Sigma$ that includes a primitive type to be interpreted as the
approximation object $\mathbb{A}$ and primitive operations to be interpreted as the monoid operations on this
object.

TODO: describe the syntactic CBN translation; show an example on the query example.

### 2025-07-10 main/models.tex
Moggi.
[ tex ] write up the semantics and some examples
[ tex ] fix bib

\begin{enumerate}
\item The category $\LatGal$ has bounded lattices as objects and
  Galois connections as morphisms, with the right adjunct going in
  the ``forward'' direction. The category $\Fam(\LatGal)$ is our
  preferred model for total functions with Galois sliicing. We
  explore some specific examples in this category in
  \secref{semantic-gps}.
\item The category $\MeetSLat$ has meet-semilattices with top as
  objects and monotone finite meet preserving functions as
  morphisms. The category $\Fam(\MeetSLat)$ provides a model of
  total functions with ``forward derivatives'' only.
\item The category $\JoinSLat$ has join-semilattices with bottom as
  objects and monotone finite join preserving functions as
  morphisms. The category $\Fam(\JoinSLat^\op)$ provides a model of
  total functions with ``backwards derivatives'' only.
\end{enumerate}

@@

Despite its similarity, this proposition has a lesser status than
\propref{embed-manifolds} because it is not clear that the category
$\Lposet$ (or $\CM$) is a canonical definition of approximable sets
and functions with approximation derivatives, as we discussed at the
end of \secref{stab}.
% We could restrict $\Lposet$ to morphisms that preserve maximal
% elements and take L-posets $X$ to
% $(\mathrm{Max}(M), \lambda x. \downset{x})$.
Our current hypothesis is that $\Fam(\LatGal)$, where values and their
approximations are separated by construction, is the natural model of
semantic \GPS in a total setting. We now investigate some categorical
properties of this category, with a view to modelling a higher-order
total programming language in \secref{language}.

@@

\subsubsection{Unapproximated Functions}

@@

\subsubsection{Lifting Monad}

@@

\subsubsection{An Approximation Object and the Tagging Monad}

@@

\subsubsection{Approximating Numbers by Intervals}

So far, the approximation lattices we have looked at in
$\Fam(\LatGal)$ have only consisted of those constructed from finite
products and lifting, and only track binary usage/non-usage
information. \exref{intervals-and-maxima-elements} shows how we can go
beyond this to get more ``quantitative'' approximation
information. Let the object of reals with interval approximations in
$\Fam(\LatGal)$ be
$(\mathbb{R}, \lambda x.\,\{[l,u] \mid l \leq x \leq u\} \cup
\{\bot\})$ where the lattices of intervals are reverse ordered by
inclusion with $\bot$ at the bottom. Then, following the examples in
\exref{intervals-and-maxima-elements} and \exref{stable-functions}, we
can define addition, negation, and scaling by $r \geq 0$:
\begin{displaymath}
  \begin{array}{lcl}
    \mathrm{add}&=&(\begin{array}[t]{@{}l}
      \lambda (x_1,x_2).\,x_1+x_2, \\
      \lambda (x_1,x_2)\,([l_1,u_1],[l_2,u_2]).\,[(l_1+x_2) \sqcap (l_2+x_1),(u_1+x_2) \sqcup (u_2+x_1)] \\
      \lambda (x_1,x_2)\,[l,u].\,([l-x_2,u-x_2],[l-x_1,u-x_1]))
    \end{array}\\
    \mathrm{neg}&=&(\lambda x.\,-x, \lambda x\,[l,u].\,[-u,-l], \lambda x\,[l,u].\,[-u,-l]) \\
    \mathrm{scale}(r)&=&(\lambda x.\,rx, \lambda x\,[l,u].\,[rl,ru], \lambda x\,[l,u].\,\mathrm{if}\,r=0\,\mathrm{then}\,\bot\,\mathrm{else}\,[\frac{l}{r},\frac{u}{r}])
  \end{array}
\end{displaymath}
Scaling by negative numbers is also possible with swapping of lower
and upper bounds, as is multiplication.

>>>>

\begin{enumerate}
\item The category $\LatGal$ has bounded lattices as objects and
  Galois connections as morphisms, with the right adjoint going in the
  ``forward'' direction. The category $\Fam(\LatGal)$ is our preferred
  model for total functions with Galois sliicing. We explore some
  specific examples in this category in \secref{semantic-gps}.
\item The category $\MeetSLat$ has meet-semilattices with top as
  objects and monotone finite meet preserving functions as
  morphisms. The category $\Fam(\MeetSLat)$ provides a model of
  total functions with ``forward derivatives'' only.
\item The category $\JoinSLat$ has join-semilattices with bottom as
  objects and monotone finite join preserving functions as
  morphisms. The category $\Fam(\JoinSLat^\op)$ provides a model of
  total functions with ``backwards derivatives'' only.
\end{enumerate}

@@

Despite its similarity, this proposition has a lesser status than
\propref{embed-manifolds} because it is not clear that the category
$\Lposet$ (or $\CM$) is a canonical definition of approximable sets
and functions with approximation derivatives, as we discussed at the
end of \secref{stab}.
% We could restrict $\Lposet$ to morphisms that preserve maximal
% elements and take L-posets $X$ to
% $(\mathrm{Max}(M), \lambda x. \downset{x})$.
Our working hypothesis is that $\Fam(\LatGal)$, where values and their
approximations are separated by construction, is a natural model of
semantic \GPS in a total setting, though we note some shortcomings in
\remref{further-structure}. We now investigate some categorical
properties of this category, with a view to modelling a higher-order
total programming language in \secref{language}.

@@

\subsubsection{Unapproximated Functions}
\label{sec:unapproximated-functions}

@@

\subsubsection{Lifting Monad}
\label{sec:models-of-total-gps:lifting}

@@

\subsubsection{An Approximation Object and the Tagging Monad}
\label{sec:tagging-monad}

@@

\subsubsection{Approximating Numbers by Intervals}
\label{sec:interval-approx}

So far, the approximation lattices we have looked at in
$\Fam(\LatGal)$ have only consisted of those constructed from finite
products and lifting, and only track binary usage/non-usage
information. \exref{intervals-and-maxima-elements} shows how we can go
beyond this to get more ``quantitative'' approximation
information. Let the object of reals with interval approximations in
$\Fam(\LatGal)$ be
$\mathbb{R}_{\mathit{intv}} = (\mathbb{R}, \lambda x.\,\{[l,u] \mid l
\leq x \leq u\} \cup \{\bot\})$ where the lattices of intervals are
reverse ordered by inclusion with $\bot$ at the bottom. Then,
following the examples in \exref{intervals-and-maxima-elements} and
\exref{stable-functions}, we can define addition, negation, and
scaling by $r \geq 0$:
\begin{displaymath}
  \begin{array}{lcl}
    \mathrm{add}&=&(\begin{array}[t]{@{}l}
      \lambda (x_1,x_2).\,x_1+x_2, \\
      \lambda (x_1,x_2)\,([l_1,u_1],[l_2,u_2]).\,[(l_1+x_2) \sqcap (l_2+x_1),(u_1+x_2) \sqcup (u_2+x_1)] \\
      \lambda (x_1,x_2)\,[l,u].\,([l-x_2,u-x_2],[l-x_1,u-x_1]))
    \end{array}\\
    \mathrm{neg}&=&(\lambda x.\,-x, \lambda x\,[l,u].\,[-u,-l], \lambda x\,[l,u].\,[-u,-l]) \\
    \mathrm{scale}(r)&=&(\lambda x.\,rx, \lambda x\,[l,u].\,[rl,ru], \lambda x\,[l,u].\,\mathrm{if}\,r=0\,\mathrm{then}\,\bot\,\mathrm{else}\,[\frac{l}{r},\frac{u}{r}])
  \end{array}
\end{displaymath}
(we only define the forward and backward maps on intervals, their
behaviour on $\bot$ is determined.)  Scaling by negative numbers is
also possible with swapping of lower and upper bounds, as is
multiplication.
